---
title: "RMarkDown"
output: pdf_document
date: "2024-09-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(caret) 
```

# Introduction
In this project, we build a movie recommendation system using the MovieLens dataset. The goal is to predict user ratings for movies they haven’t seen based on patterns in their past ratings.

# Data Loading

```{r}
# Load the data
ratings <- read.csv("ratings.csv")
movies <- read.csv("movies.csv")

# Merge the ratings and movies datasets on the movieId column
merged_data <- merge(ratings, movies, by = "movieId")

# Check the structure and the first few rows of the merged dataset
str(merged_data)
head(merged_data)
```

# Data Preparation

We split the data into training and validation sets to avoid overfitting and ensure that the model generalizes well.
```{r}
# Set seed for reproducibility
set.seed(1)

# Split the data into edx (90%) and final_holdout_test (10%)
test_index <- createDataPartition(merged_data$rating, p = 0.1, list = FALSE)
edx <- merged_data[-test_index, ]
final_holdout_test <- merged_data[test_index, ]

# Ensure final_holdout_test has only users and movies that are also in edx
final_holdout_test <- final_holdout_test %>%
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")
```

# Baseline Model: Just the Average Rating

We start by calculating the RMSE for a simple baseline model that uses the average movie rating to predict all user ratings.

```{r}
# Baseline Model: Just the average rating
mu <- mean(edx$rating)

# RMSE calculation function
rmse <- function(true_ratings, predicted_ratings) {
  sqrt(mean((true_ratings - predicted_ratings)^2))
}

baseline_rmse <- rmse(final_holdout_test$rating, mu)
cat("Baseline RMSE:", baseline_rmse, "\n")
```

# Movie Effect Model

The next model incorporates the effect of movies by adjusting for the average rating of each movie.
```{r}
# Movie Effect Model
movie_avgs <- edx %>%
  group_by(movieId) %>%
  summarize(b_i = mean(rating - mu))

predicted_ratings_movie <- final_holdout_test %>%
  left_join(movie_avgs, by='movieId') %>%
  mutate(pred = mu + b_i) %>%
  pull(pred)

movie_effect_rmse <- rmse(final_holdout_test$rating, predicted_ratings_movie)
cat("Movie Effect Model RMSE:", movie_effect_rmse, "\n")
```

# Movie + User Effect Model

We further refine the model by incorporating user-specific effects in addition to the movie effects.

```{r}
# Movie + User Effect Model
user_avgs <- edx %>%
  left_join(movie_avgs, by='movieId') %>%
  group_by(userId) %>%
  summarize(b_u = mean(rating - mu - b_i))

predicted_ratings_movie_user <- final_holdout_test %>%
  left_join(movie_avgs, by='movieId') %>%
  left_join(user_avgs, by='userId') %>%
  mutate(pred = mu + b_i + b_u) %>%
  pull(pred)

movie_user_effect_rmse <- rmse(final_holdout_test$rating, predicted_ratings_movie_user)
cat("Movie + User Effects Model RMSE:", movie_user_effect_rmse, "\n")
```

# Regularized Movie + User Effect Model

To avoid overfitting, we regularize the model by adding a penalty term to both the movie and user effects.
```{r}
# Regularized Movie + User Effect Model
lambdas <- seq(0, 10, 0.1)
best_lambda <- 0
best_rmse <- Inf

for (l in lambdas) {
  movie_reg_avgs <- edx %>%
    group_by(movieId) %>%
    summarize(b_i = sum(rating - mu) / (n() + l))
  
  user_reg_avgs <- edx %>%
    left_join(movie_reg_avgs, by='movieId') %>%
    group_by(userId) %>%
    summarize(b_u = sum(rating - b_i - mu) / (n() + l))
  
  predicted_ratings_reg <- final_holdout_test %>%
    left_join(movie_reg_avgs, by='movieId') %>%
    left_join(user_reg_avgs, by='userId') %>%
    mutate(pred = mu + b_i + b_u) %>%
    pull(pred)
  
  model_rmse <- rmse(final_holdout_test$rating, predicted_ratings_reg)
  
  if (model_rmse < best_rmse) {
    best_rmse <- model_rmse
    best_lambda <- l
  }
}

cat("Best Regularized RMSE:", best_rmse, "\n")
cat("Best lambda:", best_lambda, "\n")
```

# Conclusion

In this project, we explored multiple models to predict movie ratings based on the MovieLens dataset. The regularized Movie + User Effects Model achieved the best RMSE score. Future improvements could include incorporating time-based effects or using matrix factorization techniques for better recommendations.
```{r}
# Final Regularized Model with the best lambda
movie_reg_avgs <- edx %>%
  group_by(movieId) %>%
  summarize(b_i = sum(rating - mu) / (n() + best_lambda))

user_reg_avgs <- edx %>%
  left_join(movie_reg_avgs, by='movieId') %>%
  group_by(userId) %>%
  summarize(b_u = sum(rating - b_i - mu) / (n() + best_lambda))

predicted_ratings_final <- final_holdout_test %>%
  left_join(movie_reg_avgs, by='movieId') %>%
  left_join(user_reg_avgs, by='userId') %>%
  mutate(pred = mu + b_i + b_u) %>%
  pull(pred)

final_rmse <- rmse(final_holdout_test$rating, predicted_ratings_final)
cat("Final Regularized Model RMSE:", final_rmse, "\n")
```

